---
title: "Regression line"
author: "Janvier Ndashimye"
date: "`r Sys.Date()`"
output: html_document
---
# Fitting regression models with lm()
## Simple linear regression
```{r}
data("women")
 fit<-lm(weight~height, data=women)
 summary(fit)
```
```{r}
women$weight
```
```{r}
fitted(fit)
```

```{r}
residuals(fit)
```


```{r}
plot(women$height,women$weight,xlab="Height(ininches)",ylab="Weight
 (inpounds)")
 abline(fit)
```
The plot suggests that you might be able to improve on the prediction by using a line
 with one bend. WEcanfit thepolynomial regression. #### Polynomial regression
```{r}
fit2<- lm(weight~height+I(height^2), data=women)
 summary(fit2)
```
```{r}
plot(women$height,women$weight,xlab="Height (in inches)",ylab="Weight
 (in pounds)")
 lines(women$height,fitted(fit2))
```
#Multiple linear regression
 
 When we have more than one predictor variable, simple linear regression becomes
 multiple linear regression, and the analysis grows more involved.
 
#HW:Find a dataset of whichyou canfit multiple linear regression and interpret
 your results
#Regression diagnostics
 In this section you want to know if the model you have applied is appropriate. The
 most commonapproachistoapplythe plot() function to the object returned by the
 lm(). This produces four graphs that are useful for evaluating the model fit.
 
```{r}
fit<- lm(weight~height,data=women)
 par(mfrow=c(2,2))
 plot(fit)
```
Normality: the residual plot should be normally distributed with mean 0. The
 normal QQplot(upper right) is probability plot of the standardized residuals
 against the value that would be expected under normality. If you have met the the
 normality assumption, the points on this graph should fall on the straight 45 degree
 line. Because they don’t, you have clearly violated the normality assumption
 
 Independence: Judge based on how data were collected
 Linearity: If the dependent variable is linear related to the independent variables.
 there should be no systematic relationship between the residuals and the predicted
 (That is fitted) values. In the residuals vs. fitted graph (upper left), you see clear
 evidence of a curved relationship, which suggests that you may want to add a
 quadratic term to the regression.
 The other graph helps to identify outliers, high laverage points, influential
 observations
 
 Homoscedasticity: If you have met the constant variance assumption, the point in
 the scale-location graph(bottom left) should be a random band around a horizontal
 line. you seem to meet this assumption.
 
```{r}
 fit2<- lm(weight~height+I(height^2), data=women)
 par(mfrow=c(2,2))
 plot(fit2)
```
```{r}
newfit<-lm(weight~height+I(height^2),data=women[-c(13,15),])
 par(mfrow=c(2,2))
 plot(newfit)
```
#Outliers
 The car package also provides a statistical test outliers. the function is
 “outlierTest()”
 
#Corrective measures
 Whatdoyoudoifyouidentify problems? They are four approachesto dealing with
#violations of regression assumptions:
###Deleting observations ( influentials observation like outliers)
###Transforming variables ( transform response like y^(r))
###Adding or deleting variables ( sometimes to deal with multicolinearity)
###Using another regression approach (Like robust regression, etc…)

#Selecting the best regression model

 Shoud you include all variables under study? should you add polynomial or
 interactions terms to improve the fit? you make a decision based on predictive
 accuracy and simple and replicable model.
 HW:Readaboutvariable selection methods
 
#Exploration of mtcars data
```{r}
 plot(mpg~wt,data=mtcars)
 model<- lm(mpg~wt, data=mtcars)
 #plot(model)
 abline(model)
```
```{r}
summary(model)
```
```{r}
predict(model,newdata=data.frame(wt=6))
```
#Simulation in R
 Simulation is an important (and big) topic for both statistics and for a variety of
 other areas where there is a need to introduce randomness. Sometimes you want to
 implement a statistical procedure that requires random number generation or
 sample (i.e. Markov chain Monte Carlo, the bootstrap, random forests, bagging) and
 sometimes you want tosimulate a system andrandom numbergenerators can be
 used to model randominputs
 
#Generation of random numbers
 Wecansimulate from probability distribution
 dnorm(x, mean= 0, sd=1,log = FALSE)
 pnorm(q, mean =0, sd =1,lower.tail = TRUE, log.p = FALSE)
 qnorm(p, mean =0, sd =1,lower.tail = TRUE, log.p = FALSE)
 This gives a normal probability plot, The points in this plot will lie approximately on
 a straight line if the distribution is normal.
 rnorm(n, mean =0, sd= 1)orrnorm() Thisgenerate random numberfrom standard
 normal distribution.

```{r}
 x <- rnorm(10)
 pnorm(2)
 
 x <- rnorm(10, 20, 2)
```
#Setting the random number seed
When simulating any random numbers itis essential to set the random number seed.
 Setting the random number seed with set.seed() ensures reproducibility of the
 sequence of random numbers.
```{r}
set.seed(1)
 rnorm(5)
```
#Simulaton a linear model
```{r}
 ## Always set your seed!
 set.seed(20)
 ## Simulate predictor variable
 x <- rnorm(100)
 ## Simulate the error term
 e <- rnorm(100, 0, 2)
 ## Compute the outcome via the model
 y <- 0.5- 2 * x + e
 plot(x, y)
```
```{r}
 set.seed(1)
 sample(1:10, 4)
```
```{r}
 ##Doesn'thavetobenumbers
 sample(letters,5)
 ##[1]"r""s""a""u""w"
 ##Doarandompermutation
 sample(1:10)
 ## [1]10 6 9 2 1 5 8 4 3 7
 sample(1:10)
 ## [1] 510 2 8 6 1 4 3 9 7
 ##Samplew/replacement
 sample(1:10,replace=TRUE)
```
#Explorethemodelrpackage.
## DataVisualization
```{r}
# Define vectors
country <- c("Australia","Austria","Belgium","Canada",
             "Denmark","Finland","France","Germany",
             "Greece","Ireland","Italy","Japan","Netherlands",
             "NewZealand","Norway","Portugal","Spain","Sweden",
             "Switzerland","UK","USA")

Income.inequality <- c(7.0, 4.8, 4.6, 5.6, 4.3, 3.7, 5.6, 5.2, 6.2, 6.0, 6.7, 
                        3.4, 5.3, 6.8, 3.9, 8.0, 5.5, 4.0, 5.7, 7.2, 8.6)

Index.HS <- c(0.07, 0.01, -0.23, -0.07, -0.19, -0.43, 0.05, -0.06, 0.38, 0.25, 
              -0.12, -1.26, -0.51, 0.29, -0.63, 1.18, -0.30, -0.83, -0.46, 0.79, 2.02)

GDP_WB <- c(45926, 47682, 43435, 45066, 45537, 30676, 39328, 46401, 26851, 49393,
            35463, 36319, 48253, 37679, 65615, 28760, 33629, 45297, 59540, 40233, 54630)

# Create data frame
data.21 <- data.frame(country, Income.inequality, Index.HS, GDP_WB)

# Scatter plot of Income.inequality vs Index.HS
plot(data.21$Income.inequality, data.21$Index.HS,
     xlab = "Income Inequality",
     ylab = "Index HS",
     main = "Scatter plot: Income Inequality vs Index HS",
     pch = 19, col = "blue")


```
```{r}
Index_inequality.df<-data.21[c("Income.inequality","Index.HS")]
 str(Index_inequality.df)
```

```{r}
(country<-data.21[,"country"])
```
```{r}
 (country.2<-data.21["country"])
```

```{r}
 str(country)
```

```{r}
str(country.2)
```
```{r}
plot(Index_inequality.df,pch=20)
 text(Index_inequality.df,labels=country)
```
```{r}
plot(Index_inequality.df,pch=20)
 text(Index_inequality.df,labels=country,pos=1)
```

 
```{r}
plot(Index_inequality.df,pch=20)
 text(Index_inequality.df,labels=country,pos=3)
```
```{r}
plot(Index_inequality.df,pch=20)
 text(Index_inequality.df,labels=country,pos=4)
```
```{r}
plot(Index_inequality.df,pch=20)
 text(Index_inequality.df,labels=country,pos=4,cex=0.8)
```
```{r}
which(country %in% c("Austria","Denmark","Germany","Netherland"))
 ## [1] 2 5 8 13
 text.left<-which(country %in% c("Austria","Denmark","Germany","Netherla
 nd"))
 text.left
 ## [1] 2 5 8 13
 text.right<-setdiff(1:nrow(data.21),text.left)
 text.right
 ## [1] 1 3 4 6 7 9 10 11 12 14 15 16 17 18 19 20 21
 pos.text<-ifelse(1:nrow(data.21)%in% text.left,2,4)
 plot(Index_inequality.df,pch=20,col="red",xlim=c(3,9),ylim=c(-1.5,2.5))
 text(Index_inequality.df,labels=country,pos=pos.text,cex=0.8)

```
```{r}
# Identify Germany index
text.up <- which(country %in% "Germany")  # Germany is at position 8

# Example: specify other points for text positioning
text.left <- setdiff(c(2,5,13), text.up)  # manually specify positions if needed

# Set positions: 3=above, 2=left, 4=right
pos.text <- ifelse(1:nrow(data.21) %in% text.up, 3,
                   ifelse(1:nrow(data.21) %in% text.left, 2, 4))

# Scatter plot
plot(data.21$Income.inequality, data.21$Index.HS,
     pch = 20, col = "red", xlim = c(3, 9), ylim = c(-1.5, 2.5),
     ann = FALSE)

# Add country names
text(data.21$Income.inequality, data.21$Index.HS,
     labels = country, pos = pos.text, cex = 0.8)

# Add titles and axis labels
main.title <- "Income inequality vs Index of Health and Social problems"
x.lab <- "Income inequality (5th Ratio)"
y.lab <- "Index of Health and Social problems"

title(main = main.title, xlab = x.lab, ylab = y.lab)


```

```{r}
plot(Index_inequality.df,pch=20,col="red",xlim=c(3,9),ylim=c(-1.5,2.5),
 ann=FALSE)
 text(Index_inequality.df,labels = country,pos=pos.text,cex=0.8)
 main.title<-"Income inequality vs Index of Health and Social problems"
 x.lab<-"Income inequality (5th Ratio)"
 y.lab<-"Index of Health and Social problems"
 title(main=main.title,xlab=x.lab,ylab=y.lab)
 mtext(c("Better","Worse"),side=2,at=c(-1.8,2.8),las=1)

```
```{r}
# Scatter plot
plot(data.21$Income.inequality, data.21$Index.HS,
     pch = 20, col = "red",
     xlim = c(3, 9), ylim = c(-1.5, 2.5),
     ann = FALSE)

# Add country labels
text(data.21$Income.inequality, data.21$Index.HS,
     labels = country, pos = pos.text, cex = 0.8)

# Titles and axis labels
main.title <- "Income inequality vs Index of Health and Social problems"
x.lab <- "Income inequality (5th Ratio)"
y.lab <- "Index of Health and Social problems"
title(main = main.title, xlab = x.lab, ylab = y.lab)

# Add margin text on the y-axis
mtext(c("Better", "Worse"), side = 2, at = c(-1.5, 2.3), las = 1)

# Compute and display correlation
r_value <- cor(data.21$Income.inequality, data.21$Index.HS)
text(x = 5, y = 1.5, labels = paste("r =", round(r_value, 2)))

```
```{r}
# Scatter plot
plot(data.21$Income.inequality, data.21$Index.HS,
     pch = 20, col = "red",
     xlim = c(3, 9), ylim = c(-1.5, 2.5),
     ann = FALSE)

# Add country labels
text(data.21$Income.inequality, data.21$Index.HS,
     labels = country, pos = pos.text, cex = 0.8)

# Titles and labels
main.title <- "Income inequality vs Index of Health and Social problems"
x.lab <- "Income inequality (5th Ratio)"
y.lab <- "Index of Health and Social problems"
title(main = main.title, xlab = x.lab, ylab = y.lab)

# Add "Better"/"Worse" text on y-axis
mtext(c("Better", "Worse"), side = 2, at = c(-1.5, 2.3), las = 1)

# Compute and display correlation coefficient
r_value <- cor(data.21$Income.inequality, data.21$Index.HS)
text(x = 5, y = 1.5, labels = paste("r =", round(r_value, 2)))

# Fit linear regression and add line
lm.ineq <- lm(Index.HS ~ Income.inequality, data = data.21)
abline(lm.ineq, col = "blue", lwd = 2)


```
```{r}
rm(list=ls(all=TRUE))
 ###set.seed(1002)
 n <-30 ## number of observations
 M <- 1 ## number of pathways
 p<-5
 z <- matrix(runif(n * p, 0, 1), nrow=n, ncol=p)
 x <- 3*cos(z[, 1]) + 2*rnorm(n)
 x<-as.matrix(x)
 beta.true <- rep(1,ncol(x))
 ## pathway-response function
 hfun1 <- function(zvec) (10*cos(zvec[1])- 15*(zvec[2])^2+10*exp(-zvec
 [3])*zvec[4]-8*sin(zvec[5])*cos(zvec[3])+20*(zvec[1]*zvec[5]))
 h1 <- apply(z, 1, hfun1) ## only depends on z1,z2,z3,z4,z5
 eps <- rnorm(n)
 y <-x * beta.true + h1+ eps
```
 
 
