---
title: "Data management"
author: "Janvier Ndashimye"
date: "`17/10/2025`"
output: html_document
---
#Data Visualization and Data management

##Test to treat data
```{r}
data1 <- read.table(text = "
Sept_2022 Oct_2022 Nov_2022 Dec_2022 Jan_2023 Feb_2023 Mar_2023 Apr_2023 May_2023 June_2023
1 4 6 30 140 67 47 23 39 134 80
2 10 2 29 164 72 81 28 63 211 132
", header = TRUE)


# Open a fresh plot window (important if RStudio messed up the last one)
          
dev.new()        # opens a new window

# Draw bar plot
barplot(as.matrix(data1),
        main = "Confirmed Cases Vs Gender",
        beside = TRUE,
        col = c("gold3", "red"))

# Add legend
legend("topright",
       c("Male", "Female"),
       cex = 0.8,
       bty = "n",
       fill = c("gold3", "red"))


```
```{r}
data3 <- read.table(text = "
sept_2022 Oct_2022 Nov_2022 Dec_2022 Jan_2023 Feb_2023 Mar_2023 Apr_2023 May_2023 June_2023
1 867 582 1802 12003 3059 9205 5346 7269 5933 3245
2 842 436 1805 11785 3508 10908 4902 7883 5901 4646
", header = TRUE)

barplot(as.matrix(data3),
        main = "COVID-19 Tests Vs Gender",
        beside = TRUE,
        col = c("blue", "purple"))

legend("topleft",
       legend = c("Male", "Female"),
       cex = 1.3,
       bty = "n",
       fill = c("blue", "purple"))


```
```{r}
data2 <- read.table(text = "
Sept_2022 Oct_2022 Nov_2022 Dec_2022 Jan_2023 Feb_2023 Mar_2023 Apr_2023 May_2023 June_2023
1 0 0 1 9 8 7 1 4 11 22
2 11 3 41 242 104 78 44 86 244 147
3 3 5 17 53 27 43 6 12 90 43
", header = TRUE)

# Draw barplot
barplot(as.matrix(data2),
        main = "Confirmed Cases Vs Age Categories",
        beside = TRUE,
        col = c("green", "blue", "purple"))

# Add legend
legend("topleft",
       legend = c("Less than 15 years", "15-49 years", "Above 50"),
       cex = 1.3,
       bty = "n",
       fill = c("green", "blue", "purple"))

```

```{r}
months <- c("Sept_2022", "Oct_2022", "Nov_2022", "Dec_2022", 
            "Jan_2023", "Feb_2023", "Mar_2023", "Apr_2023", 
            "May_2023", "June_2023")

Confirmed_covid <- c(14, 8, 59, 304, 139, 128, 51, 102, 345, 212)

plot(Confirmed_covid, 
     type = "l",
     pch = 21,
     col = "red",
     ylim = c(0, 350),
     xaxt = "n",
     ylab = "Number of patients",
     main = "Confirmed Cases Vs Treatment")

treatment <- c(12, 0, 26, 217, 104, 57, 12, 47, 23, 20)

lines(treatment, col = "blue", lty = 2)

legend("topleft",
       legend = c("Received treatment", "Confirmed"),
       cex = 0.8,
       bty = "n",
       fill = c("blue", "red"))

axis(1, 
     at = 1:10, 
     labels = months,
     las = 2)  # makes labels vertical for readability

```

```{r}
months <- c("Sept_2022", "Oct_2022", "Nov_2022", "Dec_2022", 
            "Jan_2023", "Feb_2023", "Mar_2023", "Apr_2023", 
            "May_2023", "June_2023")

number_of_test <- c(1709, 1018, 3607, 23788, 6567, 20113, 10248, 15152, 11834, 7891)

plot(number_of_test, 
     type = "o",              # line with points
     col = "blue", 
     ylim = c(0, 24000),
     xaxt = "n", 
     ylab = "Number of Tests",
     main = "COVID-19 Tests")

axis(1, 
     at = 1:10, 
     labels = months,
     las = 2)   # vertical labels for better readability

```
#ggplot2

 ggplot2 provides a set of tools that allows you to visualize complex data sets in a
 newcreative way-some work needsomepackagestoget done.-some ofthe graph
 are created using R’s base graphics system
 
#Library ggplot2
```{r}
#install.packages("ggplot2")
 library(ggplot2)
```
#let us explore the data mcars
```{r}
data(mtcars)

#dotchart(mtcars$mpg,labels=row.names(mtcars),ce#x = 0.,main="miles per
 #Gallon of car #model",xlab = "MPG")

 table(mtcars$cyl)

```
```{r}
 #barplot(mtcars$cyl)
barplot(table(mtcars$cyl),
        main = "Distribution of Car Cylinder Counts",
        xlab = "Number of Cylinders",
        ylab = "Number of Cars",
        col = "lightblue")

```
```{r}
barplot(table(mtcars$cyl),
        main = "Distribution of Car Cylinder Counts",
        xlab = "Number of Cylinders",
        ylab = "Number of Cars",
        horiz = TRUE,
        col = "lightgreen")

```
```{r}
plot(mtcars$mpg ~ mtcars$wt,
     main = "Automobile Data",
     xlab = "Weight",
     ylab = "Miles per Gallon",
     pch = 19,        # solid points for better visibility
     col = "blue")

```
```{r}
# First plot
plot(mtcars$mpg ~ mtcars$wt,
     main = "Automobile Data",
     xlab = "Weight",
     ylab = "Miles per Gallon",
     pch = 19,         # solid points
     col = "blue")

# Second plot with different point type
plot(mtcars$mpg ~ mtcars$wt,
     main = "Automobile Data",
     xlab = "Weight",
     ylab = "Miles per Gallon",
     pch = 17,         # triangle points
     col = "red")

```
```{r}
plot(mtcars$mpg ~ mtcars$wt,
     main = "Automobile Data",
     xlab = "Weight",
     ylab = "Miles per Gallon",
     pch = 21,       # filled circle with border
     col = "blue",   # border color
     bg = "lightblue")  # fill color for points

```
```{r}
plot(mtcars$mpg ~ mtcars$wt,
     type = "l",        # line plot
     lty = 2,           # dashed line
     lwd = 2,           # line width
     col = "red",
     main = "Automobile Data",
     xlab = "Weight",
     ylab = "Miles per Gallon")

```
 in ggplot2, plots are created by chaining together function using (+)sign. Each function
 modify the
 
##plot created up to that point
```{r}
library(ggplot2)
 ggplot(data=mtcars, aes(x=wt,y=mpg))+geom_point()+
 labs(title = "Automobile Data",x="Weight",y="Miles per Galon")
```
```{r}
library(ggplot2)

ggplot(data = mtcars, aes(x = wt, y = mpg)) +
  geom_point(pch = 17, color = "blue", size = 1) +
  geom_smooth(method = "lm", color = "green") +
  labs(title = "Automobile Data",
       x = "Weight",
       y = "Miles per Gallon")

```

```{r}
library(ggplot2)

f0 <- ggplot(data = mtcars, aes(x = wt, y = mpg)) +
  geom_point(shape = 24, fill = 1, size = 1) +
  geom_smooth(method = "lm", color = "blue")

f0

```
```{r}
f1<-f0+theme_bw()
 f1
 ## `geom_smooth()` using formula = 'y ~ x'
```
```{r}
f2<-f1+
 theme(panel.grid = element_line(linetype = "dotted",color = "Black"))
 f2
 ## `geom_smooth()` using formula = 'y ~ x'
```

#Grouping displays two or more groups of observations in a single plot
```{r}
library(ggplot2)

data(mtcars)

# Convert variables to factors with labels
mtcars$am <- factor(mtcars$am, levels = c(0, 1),
                    labels = c("Automatic", "Manual"))
mtcars$vs <- factor(mtcars$vs, levels = c(0, 1),
                    labels = c("V-Engine", "Straight Engine"))
mtcars$cyl <- factor(mtcars$cyl)

# Create the ggplot
ggplot(data = mtcars, aes(x = hp, y = mpg, shape = cyl, color = cyl)) +
  geom_point(size = 3) +
  facet_grid(vs ~ am) +
  labs(title = "Automobile Data by Engine Type",
       x = "HorsePower",
       y = "Miles per Gallon")

```
```{r}
# Load necessary packages
library(lattice)
library(ggplot2)

# Load the dataset
data(singer, package = "lattice")

# Create boxplot using ggplot2
ggplot(singer, aes(x = voice.part, y = height)) +
  geom_boxplot() +
  labs(title = "Height by Voice Part",
       x = "Voice Part",
       y = "Height (inches)") +
  theme_minimal()

```
#Usethefollowingfunction:

 geom_bar()geom_boxplot()geom_density()geom_histogram()geom_hline()
 geom_jitter()geom_line()#geom_point()geom_rug()#geom_smooth()geom_text()
 geom_violin()geom_vline()##Basicdatamanagement
 
#cbindandrbind

 Whilecombiningcolumnwise,thenumberofrowsmustmatchbutrownamesare
 ignored.whencombiningrow-wise,boththenumberandthenamesofcolumns
 mustmatch.
 
```{r}
data2 <- data.frame(x = 1:3, y = c("a", "b", "c"))
str(data2)

```
```{r}
(cbind(data2,data.frame(z=3:1)))
```
```{r}
 (rbind(data2,data.frame(x=10,y="z")))
```
#CreateanotherVariable
```{r}
data_class <- read.csv("C:/Users/NDASHIJANVIER/Documents/AUCA/Big Data Analytics/R/course/Assignments/dataset/data_class.csv")

 variable.names(data_class)
```
```{r}
head(data_class,n=5)
```
```{r}
 tail(data_class)
```
```{r}
data_class[,1]
```
```{r}
 summary(data_class$WEIGHT)
```
```{r}
length(data_class$WEIGHT)
```
```{r}
data_class[,-1]
```
```{r}
attach(data_class)
 (BMI<-WEIGHT/(HEIGHT/100)^2)
```
```{r}
 (BMI<-round(WEIGHT/(HEIGHT/100)^2,digit=1))
```
```{r}
 head(cbind(data_class,BMI))
```
```{r}
tail(cbind(data_class,BMI),n=10)
```
```{r}
 detach(data_class)
```
#SummaryofBMI
```{r}
 summary(BMI)
```
```{r}
hist(data_class$HEIGHT)
```


```{r}
hist(data_class$WEIGHT,
     col = "blue",
     main = "Sample Histogram",
     xlab = "Weight",
     ylab = "Frequency",
     border = "black")

```
#Subsetting

##Useof$,[].or[[]]

 CreationofleadershipdataInterest:Howmenandwomendifferinthewaythey
 leadtheirorganizations.5questionswereaskedinthisstudy.Exampleofthe
 question:Domenandwomeninmanagementpositiondifferinthedegreetowhich
 theydefertosuperiors?1:stronglydisagree,2:disagree,3:neitheragreenor
 disagree,4:agree,5:stronglyagree
 
```{r}
# Create vectors
manager <- c(1, 2, 3, 4, 5)
country <- c("US", "US", "UK", "UK", "UK")
gender <- c("M", "F", "F", "M", "F")
age <- c(32, 45, 25, 39, 99)
q1 <- c(5, 3, 3, 3, 2)
q2 <- c(4, 5, 5, 3, 2)
q3 <- c(5, 2, 5, 4, 1)
q4 <- c(5, 5, 5, NA, 2)
q5 <- c(5, 5, 2, NA, 1)

# Create data frame
leadership <- data.frame(manager, country, gender, age, q1, q2, q3, q4, q5,
                         stringsAsFactors = FALSE)

# Check structure
str(leadership)

```
```{r}
names(leadership)
```
 Whatyoucando-combinethescoreofthefivequestions-handlethemissing
 values-createadatasetofwhatyouwant-createagegrouporagecategories-99
 indicatethevalueismissing
 
#Missingvalue
```{r}
# Replace age 99 with NA
leadership$age[leadership$age == 99] <- NA

# Create age category column
leadership$agecat <- NA

leadership$agecat[leadership$age > 75] <- "Elder"
leadership$agecat[leadership$age >= 35 & leadership$age <= 75] <- "MiddleAged"
leadership$agecat[leadership$age < 35] <- "Young"

# Check result
leadership

```
#Onewaytohandlemissingvalue

 Deletingallobservationswithmissingdata(Listwisedeletion)isoneoftheseveral
 methodsofhandlingincompletedatasets.Note:Youcanalsoreplacingthemissing
 valuebytheaverageoftheremainingdata.

```{r}
is.na(leadership[,5:9])
```
```{r}
newdata<-na.omit(leadership)
 newdata
```

# Sortingdata
```{r}
 order(leadership$age)
```
```{r}
newdata2<-leadership[order(leadership$age),]
```
#Mergingdatasets

 Tomergetwodataframehorizontally,youusemerge()function.Inmostcases,two
 dataframesarejoinedbyoneormorecommonkeyvariables.-example1:
 merge(dataframeA,dataframeB,by=“ID”)-example2:
 merge(dataframeA,dataframeB,by=c(“ID”,“country”))
 thesecondmergethetwodataframesbyIDandcountry
 Tojointwodataframe(datasets)vertically,userbind()function:Notethatthetwo
 datasetmusthavethesamevariables
 
#Subset(selectingvariables,droppingvariables,selectingobservation)
```{r}
# Extract columns 5 to 9 by index
newdata3 <- leadership[, 5:9]

# Extract specific columns by name
myvars <- c("q1", "q2", "q3", "q4", "q5")
newdata3 <- leadership[myvars]

# Remove specific columns by index
newdata4 <- leadership[, -c(1, 2)]   # removes columns 1 and 2
newdata4 <- leadership[, -(1:5)]     # removes columns 1 to 5

# Remove columns 1 and 7
newdata5 <- leadership[, -c(1, 7)]

# Access the 4th column directly
leadership[[4]]       # equivalent to leadership$age

```

```{r}
 leadership$age
```
```{r}
newdata6<-leadership[c(-1,-3),]
 newdata6<-leadership[c(2,4,5),]
 newdata7<-subset(leadership,age>=35|age<24,select=c(q1,q2,q3,q4,q5))
 attach(leadership)
```
```{r}
#newdata7<-leadership[gender="M"&age>30,]
 newdata7<-leadership[gender=="M"&age>30,]
 detach(leadership)
```
#Datamanagementwithdyplr
 Datasourcefromthepackagenycflights13andggplot2Thisdatacontainsall
 336,776flightsthatdepartedfromNewYorkcityin2013.
 
#Datamanagementwithdyplr
 Datasourcefromthepackagenycflights13andggplot2Thisdatacontainsall
 336,776flightsthatdepartedfromNewyorkcityin2013.


```{r}
#install.packages("nycflights13")
 #install.packages("tidyverse")
 library(nycflights13)

 #install.packages("dplyr")
 library(dplyr)
```
# Thefunctionfilter()
```{r}
#rm(list=ls())
 str(flights)
```
```{r}
#attach(flights)
 jan<-filter(flights,month==1)
 jan
```

```{r}
 newyear<-filter(flights,month==1,day==1)
 newyear
```

```{r}
dec25<-filter(flights,month==12,day==25)
 dec25
```
```{r}
 jan_dec<-filter(flights,month==1|month==12)
 jan_dec
```
```{r}
 nov_dec<-filter(flights,month%in%c(11,12))
 #detach(flights)
```
#Thefunctionarrange()
 Thischangetheorder
```{r}
data_10<-arrange(flights,year,month,day)
 arrange(flights,desc(arr_delay))
```
#Thefunctionselect
 This helps to selects only the variables you are interested in
```{r}
 time_var<-select(flights,year,month,day)
 select(flights,year:day)
```
```{r}
select(flights,-(year:day))
```
# Thefunctionmutate
 ThisisusefulinaddinganewVariable
```{r}
 ##createasmalldataset
 flights2<-select(flights,year:day,ends_with("delay"),distance,air_time)
 (mutate(flights2,gain=arr_delay-dep_delay,
 speed=distance/air_time*60))
```

 

 
 
 





 











 
