---
title: "Control structures"
author: "Janvier Ndashimye"
date: "17/10/2025`"
output: html_document
---
#Grouping
 Control structures in R allow you to control the flow of execution of a series of R
 expressions. Basically, control structures allow you to put some “logic” into your R
 code, rather than just always executing the same R code every time. Control
 structures allow you to respond to inputs or to features of the data and execute
 different R expressions accordingly. Commonly used control structures are • if and
 else: testing a condition and acting on it • for: execute a loop a fixed number of times
 • while: execute a loop while a condition is true • repeat: execute an infinite loop
 (must break out of it to stop) • break: break the execution of a loop • next: skip an
 interaction of a loop
 if (condition){ ## do something } ## continue with the rest of the code.
 if( condition){ do something
 } else { do something else
 }
 Youcanalso create a series of test by following the initial if with a number of else ifs
 if(condition){ do something
 }else if (condition 2){ Do something different }else{ do something different }
 
# Function on R
```{r}
 f<-function(x) x^2
 formals(f)
```
```{r}
body(f)
```
```{r}
 environment(f)
```
#ifelse(test,yes,No)
```{r}
# Create p-values
pvalues <- c(0.867, 0.0054, 0.0018, 0.1572, 0.0183, 0.5386)

# Determine significance
results <- ifelse(pvalues < 0.05, "Significant", "Not significant")
results
# [1] "Not significant" "Significant" "Significant" "Not significant" "Significant" "Not significant"

# Example of conditional assignment
x <- runif(1, 0, 10)  # random number between 0 and 10

if (x > 3) {
  y <- 10
} else {
  y <- 0
}

# Print values
x
y

```
Thevalueofyisasetdependingonwhetherx>3ornot.Thiscanalsobeachieved
 by
```{r}
 y<-if(x>3){
 10
 }else{
 0
 }
```
####For
```{r}
for (i in 1:10) {
  print(i)
}

```


```{r}
x <- c("a", "b", "c", "d")

# Access the 3rd element
x[3]
# [1] "c"

# Loop over elements safely
for (i in 1:length(x)) {
  print(x[i])
}

```
```{r}
for (i in 1:5) {
  print(1:i)
}

```

```{r}
for (i in 5:1) {
  print(1:i)
}

```
#whileLoops
 Itbeginsbytestingacondition,ifitistrue,thentheyexecutetheloopbody.oncethe
 loopbodyisexecuted,theconditionistestedagain,untiltheconditionisfalse.after
 whichtheloopexits.
```{r}
count <- 1

while (count < 10) {
  print(count)
  count <- count + 1
}

```
```{r}
z <- 5
set.seed(1)

while (z >= 3 && z <= 10) {
  coin <- rbinom(1, 1, 0.5)
  
  if (coin == 1) {
    z <- z + 1
  } else {
    z <- z - 1
  }
}

print(z)

```
#next,break
 Thisisusedtoskipaniterationofaloop
```{r}
for (i in 1:100) {
  print(1:i)
  
  if (i > 20) {
    break
  }
}

```
#Function
 functionname<-function(parameters){statementsreturn(value)}
 f1<-function(x){result<-x^2+2return(result)}
 f2<-function(x,y){result<-x2+y2-4return(result)}
```{r}
 f<-function(x,y){
 result<-x+(2*y)+3
 return(result)
 }
f(2,3)
```
 Youcanuseargs()functiontoviewtheparameternamesanddefaultvalues
#Exercises
 Makefunctionsthatcalculatesummarystatistics
 Makeafunctiontocalculatetwosamplettest
#Applyingfunctionstomatricesanddataframe.
```{r}
a<-4
 sqrt(4)
 
```
```{r}
b<-c(1,243,5.754,2.987)
 round(b)
```
```{r}
 c<-matrix(runif(12),nrow=3)
 c
```
```{r}
log(c)

```
```{r}
 mean(c)
```
 Noticethatthemeanofmatrixcresultsinascalar(0.444).themean()takethe
 averageofall12elementsinthematrix.Butwhatifyouwantthethreerowmeans
 orthefourcolumnmeans?
 Rprovidesafunction,apply()thatallowstoapplyanarbitraryfunctiontoany
 dimensionofamatrix,arrayordataframe.Theformatfortheapply()functionis
apply(x, MARGIN, FUN,…) where x stands for the data object, Margin can be 1(rows)
 and 2(columns ) ##Col/Row SumsandMeans
 
  • rowSums=apply(x, 1, sum) • rowMeans= apply(x, 1, mean)• colSums =apply(x, 2,
 sum) •colMeans= apply(x, 2, mean)
 
#lappy function
 lapply() takes three inputs:,x,a list,a function, and…, It applies to each element of the
 list and returns a new list. lapply(x,f,…). It is called function because it takes function
 as anargument. Assume wehaveadataframe df. instead of assigningthe result pf
 lappliy() to df, we will assign them to df[] to ensure we get a data frame.
 
```{r}
 fix_missing<-function(x){
 x[x==-99]<-NA
 x
 }
 function1<-function(x){
 x^2
 }
 curve(function1,-4,4)
```
 
#df[]<-lappy(df,fix_missing)
 It works for any numberofcolumns.There is no wayto accidentally miss a column
Thereisnowaytoaccidentallytreatonecolumndifferentlythananother
 Iteasytogeneralizethistechniquetoasubsetofcolumns
#df[1:5]<-lapply(df[1:5],fix_missing)
 
#sapply
 sapply()andvapply(),variantsoflapply()thatproducesvectors,matrices,and
 arraysasOutput,insteadoflistsmap(),andmapplywhichiterateovermultiple
 inputdatastructuresinparallel
 Anotherimportantfunctionwhendealingwithbigdataissplit()-tapply

#Titanicdata
```{r}
 str(Titanic)
```
```{r}
View(Titanic)
 apply(Titanic,c(1,2),sum)
```
```{r}
 options(digits=2)
 apply(Titanic,c(1,2),sum)[3:4,]
```
```{r}
apply(Titanic,c(1,4),sum)[3:4,]
```
```{r}
apply(Titanic,c(1,2,4),sum)
```
```{r}
 #apply(Titanic,c(1,2,4),sum)[3:4,]
 apply(Titanic,c(1,2,4),sum)[3:4,,]
```
```{r}
ftable(apply(Titanic,c(1,4,2),sum)[3:4,,])
```
```{r}
ftable(apply(Titanic,c(1,4,2),sum)[3:4,,])[1:2,]
```

```{r}
 ftable(apply(Titanic,c(1,4,2),sum)[3:4,,])[3:4,]
```
```{r}
 digit=2
 prop.table(ftable(apply(Titanic,c(1,4,2),sum)[3:4,,])[1:2,],margin=2)
```
```{r}
prop.table(ftable(apply(Titanic,c(1,4,2),sum)[3:4,,])[3:4,],margin=2)
```
```{r}
# Compute proportion table for Titanic dataset
mat <- matrix(
  prop.table(
    ftable(
      apply(Titanic, c(1, 4, 2), sum)[3:4, , ]
    )[1:2, ],
    margin = 2
  ),
  nrow = 2,
  dimnames = list(
    dimnames(Titanic)$Survived,
    dimnames(Titanic)$Sex
  )
)

mat

```
```{r}
# Compute column-wise proportion table for Titanic dataset
mat <- matrix(
  prop.table(
    ftable(
      apply(Titanic, c(1, 4, 2), sum)[3:4, , ]
    )[3:4, ],
    margin = 2
  ),
  nrow = 2,
  dimnames = list(
    dimnames(Titanic)$Survived[3:4],
    dimnames(Titanic)$Sex
  )
)

mat

```

```{r}
# Create dataset
Student <- c("Johnncuti", "Angelabakame", "Brucewizeye", "Alexisaganze",
             "ClaudeRukundo", "JoelKagabo", "Maryineza")
Math <- c(600, 412, 358, 495, 512, 410, 522)
Science <- c(95, 99, 80, 82, 75, 89, 77)
English <- c(25, 22, 18, 20, 29, 30, 27)

# Create data frame
roster <- data.frame(Student, Math, Science, English, stringsAsFactors = FALSE)

# Check structure
str(roster)

```
 standardize variables and obtains the performance scores because they are reported on different scale(With widely differing means and standard deviations,we need to make them comparable before we combine them.)
 
```{r}
z<-scale(roster[,2:4])
```
###performance of each students using row means and adding them to roster using cbind()
```{r}
score<-apply(z,1,mean)
 roster<-cbind(roster,score)
```
###Grades thestudents:quantile function gives the percentile rank 
###of each student's performance score check the cutoff of A
```{r}
y<-quantile(score,c(.8,.6,.4,.2))
```
```{r}
 ##createagradevariableus
 roster$grade[score>=y[1]]<-"A"
 roster$grade[score<y[1]&score>=y[2]]<-"B"
 roster$grade[score<y[2]&score>=y[3]]<-"C"
 roster$grade[score<y[3]&score>=y[4]]<-"D"
 roster$grade[score<y[4]]<-"F"
```
```{r}
 ####Dealingwithnames
 name<-strsplit((roster$Student),"")
 lastname<-sapply(name,"[",2)
 firstname<-sapply(name,"[",1)
 roster<-cbind(firstname,lastname,roster[,-1])
 roster<-roster[order(lastname,firstname),]
 roster
```

#Aggregationandreshaping

##transpose
```{r}
cars<-mtcars[1:5,1:4]
 cars
```
```{r}
 t(cars)
```
#aggregate data
 aggregate() collapse data in R using one or more by variables and a defined function
```{r}
# Set digits to display
options(digits = 3)

# Aggregate numeric columns by cyl and gear
aggdata <- aggregate(
  mtcars[, sapply(mtcars, is.numeric)],
  by = list(Cyl = mtcars$cyl, Gear = mtcars$gear),
  FUN = mean,
  na.rm = TRUE
)

aggdata

```
#Reshape
 Step 1: install the package reshape2 step 2: melt data step 3: Cast the melted data
 into any shape you desire
 During the cast, you can aggregate the data with any function you wish.
 
```{r}
 id<-c(1,1,2,2)
 time<-c(1,2,1,2)
 x1<-c(5,3,6,2)
 x2<-c(6,5,1,4)
 mydata<-data.frame(id,time,x1,x2)
 mydata
```

 
 



 
 




