---
title: "Class1-R-2025"
author: "Janvier Ndashimye"
date: "`r Sys.Date()`"
output: html_document
---

##Data structure (Vectors, Factors, Lists, Data frame, matrix and arrays)
##Vector: The foundational R data structure
```{r}
 x <- c(0.5, 0.6)
 age<-c(20,35,32,29)
 summary(age)
 
```
```{r}
 age<-c(35,24,18,24)
 mean(age)
```
```{r}
 #Logical
 x <- c(TRUE, FALSE)
 x <- c(T, F)
 
 #Character
 x<-c("a","b","c")
 class<-c("M","F","F","M")
 
 #Integer
 x<-9:29
 
 #Complex
 x<-c(1+0i,2+4i)
 
 #Matrixandoperationwithinmatrices
 x<-matrix(1:6,nrow=2,ncol=3)
 x
 
  (y<-matrix(1:6,nrow=3,ncol=2))
 
  #List
 x<-list(1,"a",TRUE,1+4i)
 x
```

```{r}
 #Factor
 x<-factor(c("yes","yes","no","yes","no"))

 #CreateavectorwithNAsinit
 x<-c(1,2,NA,10,3)

```
ReturnalogicalvectorindicatingwhichelementsareNA
```{r}
 is.na(x)

x<-c(1,2,4,"NA",5)
 bad<-is.na(x)
 print(bad)
 x[!bad]
```
WhatiftherearemultipleRobjectsandyouwanttotakethesubsetwithnomissing
valuesinanyofthoseobjects?
```{r}
 x<-c(1,2,NA,4,NA,5)
 y<-c("a","b",NA,"d",NA,"f")
 good<-complete.cases(x,y)
 good
```
 Ifcharacterispresent,inavector,Rconverteverythinginthevectortocharacter
 strings.
 Ifavectoronlycontainslogicalandnumbers,Rwillconvertthelogicaltonumbers,
 Everytruebecomesa1,andeveryFALSEbecomes0
 
```{r}
 sum(c(TRUE,TRUE,FALSE,FALSE,FALSE))
```
# Create data
 CreatedataUsingdataframeDataframeismoregeneralthanmatrixHow???
 Becausedataframecancontaindifferentmodesofdata(Numeric,characterandso
 on)SimilartowhatyoucanseeinSPSS,SAS,…
 
###Let’screateadataframe
```{r}
studentID<-c(1,2,3,4,5)
math_score<-c(12,17,10,9,NA)
gender<-c("M","F","M","M","F")
it_score<-c(13,18,11,10,19)
scoredata<-data.frame(studentID,gender,math_score,it_score)
scoredata

```
#View(scoredata)

 Createdatafromkeyboard
 Steps1.Createdataframe(ormatrix)withvariablenames2.Invokethetexteditor
 inthe data objected created at first step
 
```{r}
 data_class2<-data.frame(height=numeric(0),weight=numeric(0),bmi=numeric
 (0))
 data_class2<-edit(data_class2)
```
# DataImportation
 Rhassomefeaturesthatcanallowtoimportdatafromdifferentsources(Itcanbe
 textfile,spreadsheet,ordatabase)
 1. Datafromexcel
 2. Datastatisticalpackages(SAS,SPSS,Stata)
 3. DatafromTextfiles(ASCII,XML,Webscraping)
 4. Datafromdatabasemanagementsystems(SQL,MySQL,Oracle,Access)
 
#  HW1:ImportdatafromStatisticalpackageandfromDatabasemanagementsystems
```{r}
data_class <- read.csv("C:/Users/NDASHIJANVIER/Documents/AUCA/Big Data Analytics/R/course/Assignments/dataset/data_class.csv")

variable.names(data_class)
 
head(data_class)

```
```{r}
tail(data_class)
```

 
```{r}
data_class[10:20,]
```

```{r}
summary(data_class$WEIGHT)

```

```{r}
length(data_class$WEIGHT)
```
```{r}
data_class[,-1]
```
 IncaseyouwanttousedatasetbuiltinR
```{r}
data() #listofdatasetscurrentlyavailable
 data("airquality")
 variable.names(airquality)
```
```{r}
str(airquality)
```
Explorationdataanalysis
 Examinenumericalvariableusingcommonsummary
 
```{r}
data_class
```
```{r}
summary(data_class)
```
```{r}
str(data_class)
```
# SummaryStatistics
```{r}
summary(data_class)
```
```{r}
 apply(data_class,2,mean)
```

```{r}
apply(data_class,2,sd)

```
```{r}
 c(mean(data_class$HEIGHT),sd(data_class$HEIGHT))
```
```{r}
 c(mean(data_class$WEIGHT),sd(data_class$WEIGHT))
```
```{r}
 c(Mean=mean(data_class$HEIGHT),SD=sd(data_class$HEIGHT))

```

```{r}
c(Mean=mean(data_class$WEIGHT),SD=sd(data_class$WEIGHT))
```
#Variationandcovariationusingbothnumericalandgraphs
#Relationshipbetweenacategoricalandacontinuousvariable
```{r}
 data("iris")
 head(iris)
```
```{r}
mean(iris$Petal.Length)

mean(iris$Petal.Length[iris$Species=="setosa"])
 
mean(iris$Petal.Length[iris$Species=="versicolor"])

mean(iris$Petal.Length[iris$Species=="virginica"])
```
##shortcut
```{r}
by(iris$Petal.Length,iris$Species,mean)
```
```{r}
by(iris$Petal.Length,iris$Species,sd)
```
```{r}
by(iris$Petal.Length,iris$Species,summary)
```
#Visualize the differences in continuous variables between categories using Box-and
#hisker plot
```{r}
boxplot(iris$Sepal.Length~iris$Species,data=iris, main="Comparison")

```

##Relationship between two categorical variables
```{r}
ucba<-data.frame(UCBAdmissions)
 head(ucba)
```
```{r}
 cross<-xtabs(Freq~Gender+Admit,data=ucba)
 (cross<-xtabs(Freq~Gender+Admit,data=ucba))
```

##IstheregenderbiasinUCBgraduateadmissionprocess?
```{r}
 prop.table(cross,2)
```
## Simpson’sparadox
 Phenomenon,whereatrendthatappearsincombinedgroupsofdatadisappearsor
 reverseswhenbrokendownintogroups.
```{r}
cross2<-xtabs(Freq~Gender+Admit,data=ucba[ucba$Dept=="A",])
 prop.table(cross2,1)
```
## BarChart
```{r}
dat<-read.table(text="ProdAProdBProdCProdD
 1110506070
 2120508065",header=TRUE)
 barplot(as.matrix(dat),beside=FALSE,col=c("Red","green"))
```
#barplot(as.matrix(dat),beside=TRUE,col=c("gold3","red"))
```{r}
 dat <- read.table(text = "A B C D E F G
 1 10 80 30 90 70 60 90
 2 20 50 70 50 40 10 40
 3 60 80 80 60 60 30 160
 4 20 40 70 80 20 10 70", header = TRUE)
 barplot(as.matrix(dat))
```
#Scatter Plot
```{r}
 plot(WEIGHT~HEIGHT,data=data_class)
```
```{r}
plot(data_class$HEIGHT,data_class$WEIGHT)
```
```{r}
 plot(data_class[,1],data_class[,2])
```

```{r}
cor(data_class$HEIGHT,data_class$WEIGHT)
```
##Linear Models
```{r}
plot(WEIGHT~HEIGHT,data=data_class)
abline(lm(WEIGHT~HEIGHT,data=data_class)$coefficient)
```
```{r}
 data_lm<-lm(WEIGHT~HEIGHT,data=data_class)
 summary(data_lm)
```
```{r}
 hist(data_class$HEIGHT)
```

```{r}
hist(data_class$WEIGHT,col = "blue")
```
#Linear with other line
```{r}
 plot(WEIGHT~HEIGHT,data=data_class)
 abline(lm(WEIGHT~HEIGHT,data=data_class)$coefficient)
 lines(lowess(data_class$HEIGHT,data_class$WEIGHT),col="blue")
```

```{r}
cor(data_class$HEIGHT,data_class$WEIGHT)
 
 model1<-lm(WEIGHT~HEIGHT,data=data_class)
```
##Data management

###Create another Variable
```{r}
attach(data_class)
 (BMI<-WEIGHT/(HEIGHT/100)^2)
```
```{r}
(BMI<-round(WEIGHT/(HEIGHT/100)^2,digit=2))
```
```{r}
 head(cbind(data_class,BMI),n=5)
```
```{r}
new_data_class<-cbind(data_class,BMI)
 tail(cbind(data_class,BMI),n=10)
```

##SummaryofBMI
```{r}
summary(BMI)
```
```{r}
stem(BMI,scale=2)
```













 
 

 




 
 